Formalizing the Edmonds-Karp Algorithm

### 2016-03-02 paper/root.tex
modified FoFu

\section{A short background on the Ford-Fulkerson method}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}
A flow network is a weighted digraph where the set of vertices (nodes) $V$ is finite, and each edge $(u, v)$ has a real capacity $c (u, v) \geq 0$. There are two distinct vertices $s$ (source) and $t$ (sink) in the flow network. In this case the network is called an $s$-$t$ network. To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties.

An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions. 1) \textbf{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \textbf{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.

@@

The total flow value is increased by manipulating the flow values along different edges. However, the flow along specific edges might decrease during this process. In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network. Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network. It also and consists of edges with capacities representing how flow can be changed along the edges of the flow network.

@@

The intuition behind the above definition is based on the definition of the residual graph. The augment procedure increases the flow on $(u, v)$ by $f ( u, v )$ but decreases it by $f (v, u)$. This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network. In each iteration of the algorithm we replace the current flow with the result of the augment function, and the algorithm terminates if no more augmenting paths can be found in the residual graph. According to the Ford-Fulkerson theorem, the flow computed in this states is indeed the maximum flow in the flow network. The theorem states that following statements are equivalent:

@@

The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found. For the networks with irrational edge capacities the algorithm might even fail to terminate. In practice, the maximum-flow problem often arises with integral capacities. Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer. If $f^*$ denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration. Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in $O (E + V)$.

The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph. There are several implementations with different execution times. For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time. The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph. Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and it always terminates. Which finally gave the polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.

>>>>

\section{A short background on the Ford-Fulkerson method}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}
A flow network is a weighted digraph where the set of vertices (nodes) $V$ is finite, and each edge $(u, v)$ has a real capacity $c (u, v) \geq 0$. There are two distinct vertices $s$ (source) and $t$ (sink) in the flow network. In this case the network is called an $s$-$t$ network. To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.

An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions. 1) \emph{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.

@@

The total flow value is increased by manipulating the flow values along different edges. However, the flow along specific edges might decrease during this process. In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network. Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network and consists of edges with capacities representing how flow can be changed along the edges of the flow network.

@@

The intuition behind the above definition is based on the definition of the residual graph. The augment procedure increases the flow on $(u, v)$ by $f ( u, v )$ but decreases it by $f (v, u)$. This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.

In each iteration of the algorithm we replace the current flow with the result of the augment function--- another flow on the network but with greater value. The algorithm terminates if no more augmenting paths can be found in the residual graph. According to the Ford-Fulkerson theorem, the flow computed after termination is indeed the maximum flow in the flow network. According to the theorem, the following statements are equivalent:

@@

The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found. For the networks with irrational edge capacities the algorithm might even fail to terminate~\cite{Zwick95}. In practice, the maximum-flow problem often arises with integral capacities. Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer. If $f^*$ denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration. Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in $O (E + V)$.

The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph. There are several implementations with different execution times. For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time. The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph. Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates. Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.
